Thesis Title

### 2022-03-02 chapters/introduction.tex

As stated previously (see \secref{sec:genomics_approach} and \secref{ssec:metagenomics}), one of the biggest challenges when dealing with genomic, and in particular metagenomic, data is the lack of golden standards. This is also applicable to the bioinformatic analysis, required due to the amount of data produced by genomic sequencing technologies. This is currently one of the bottleneck in the deployment of sequencing technology in clinical microbiology as there's no standard in how to deal with the increasing amount of data produced in a fit-for-purpose manner \citep{carrico_primer_2018}.

@@

>>>>

As stated previously (see \secref{sec:genomics_approach} and \secref{ssec:metagenomics}), one of the biggest challenges when dealing with genomic, and in particular metagenomic, data is the lack of golden standards. This is also applicable to the bioinformatic analysis, required due to the amount of data produced by genomic sequencing technologies. This is currently one of the bottleneck in the deployment of sequencing technology in clinical microbiology as there's no standard in how to deal with the increasing amount of data produced in a fit-for-purpose manner \citep{carrico_primer_2018}. Bioinformatics is an interdisciplinary research field that applies methodologies from computer science, applied mathematics and statistics to the study of biological phenomena\citep{carrico_primer_2018}. With the widespread use and continuous development of sequencing technologies, bioinformatics has become a cornerstone in modern clinical microbiology.

@@

\subsection{The FASTQ file}

All sequencing technologies, regardless of generation, produce data in the the same standard file format: the FASTQ, a text-based format for storing both a biological sequence (usually nucleotide sequence) and its corresponding quality scores \citep{cock_sanger_2010}. Originally developed at the Wellcome Trust Sanger Institute, the FASTQ has emerged as a common file format for sharing sequencing read data (see \ref{fig:figure3}). The FASTQ can be considered as an extension of the ‘FASTA sequence file format’, originally invented by \cite{pearson_improved_1988}, which includes just the sequence information. A FASTQ file normally uses four lines per sequence:

\begin{itemize}
    \item \textbf{Line 1} begins with a '@' character and is followed by a sequence identifier and an optional description;
    \item \textbf{Line 2} is the raw sequence letters;
    \item \textbf{Line 3} begins with a '+' character and is optionally followed by the same sequence identifier (and any description) again;
    \item \textbf{Line 4} encodes the quality values for the sequence in Line 2, and must contain the same number of symbols as letters in the sequence.
\end{itemize}

In FASTQ both the sequence letter and quality score are each encoded with a single ASCII character for brevity. The quality of a sequence in a FASTQ file is represented by a quality value Q is an integer mapping of p, where p is the probability that the corresponding base call is incorrect (see \ref{tab:phred_quality}). This is called the Phred score \citep{ewing_base-calling_1998} and is defined by the following equation:

\begin{equation}
\centering
\mathbf{Q}\textsc{phred} = - 10 \times \log \mathbf{P}
\end{equation}

The Phred quality scores \mathbf{Q} is defined as a property which is logarithmically related to the base-calling error probability \mathbf{P}.

\begin{table*}[h!]
\begin{tabular}{@{}lll@{}}

\caption{\textbf{Phred quality scores are logarithmically linked to error probabilities.}}\label{tab:phred_quality}

\toprule
Phred Quality Score & Probability of incorrect base call & Base call accuracy \\ 
\midrule
10                  & 1 in 10                            & 90\%               \\
20                  & 1 in 100                           & 99\%               \\
30                  & 1 in 1000                          & 99.9\%             \\
40                  & 1 in 10,000                        & 99.99\%            \\
50                  & 1 in 100,000                       & 99.999\%           \\
60                  & 1 in 1,000,000                     & 99.9999\%          \\ \bottomrule
\end{tabular}
\end{table*}

Since their introduction, Phred scores have become the \textit{de facto} standard for representing sequencing read base qualities \citep{cock_sanger_2010}. Despite this convention, the encoding of the Phread score can vary when it is translated to its ASCII representation in the FASTQ file format. For example, the Sanger FASTQ files use ASCII 33–126 to encode Phred qualities from 0 to 93 (i.e. Phred scores with an ASCII offset of 33). A full list of available encoding is available in \ref{}. %ADD FIGURE!

\subsubsection{FASTQ file simulation}

\subsubsection{FASTQ quality assessment and quality control}

Quality assessment and control is a basal step to any analysis, and aims to remove and/or filter low quality and low complexity reads, trim adapters, and remove host sequences from the samples’ raw data. There are many tools available but the most commonly used are FastQC\footnote{\url{https://www.bioinformatics.babraham.ac.uk/projects/fastqc/}} (Babraham Bioinformatics) for quality control, followed by Trimmomatic \citep{bolger_trimmomatic_2014}, Cutadapt \citep{martin_cutadapt_2011} or fastp \citep{chen_fastp_2018} to trim and/or filter adaptors, low quality and low complexity sequences. For long-read sequencing, tools like NanoPlot and NanoStats \citep{de_coster_nanopack_2018}, and Filtlong\footnote{\url{https://github.com/rrwick/Filtlong/}} can perform basic quality assessment and control, adapter trimming and low quality trimming respectively.

The removal of host sequences is usually done through mapping. Bowtie2 \citep{langmead_fast_2012} is usually the go-tool for this process, specially as pre-built indexes of the human genome are available on-line\footnote{\url{https://support.illumina.com/sequencing/sequencing\_software/igenome.html}}, useful when processing clinical metagenomic samples. Minimap2 \citep{li_minimap2_2018} has recently gained a lot of traction due to its minimizers approach that reduce the mapping time to a fraction of the one needed for Bowtie2 with comparable accuracy. Both tools can handle reads from all generations of sequencing instruments. The same approach can be used to remove contaminants present in the negative control by using it as reference for mapping.

\subsection{From reads to genomes}

The term ‘draft genome’ is commonly used because the new technologies that use short reads do not generate a single closed genome but rather a series of sequences (contigs) that may cover up to 95\% to 9\% of the strain genome.
